Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2015-11-20 pldi-2016/paper.tex
Fix typo and make things fit

Applications for social network clients, finding tomorrow's weather or searching train schedules
all communicate with external services. Increasingly these services provide REST-based 
end-points that return data as CSV, XML or JSON. Despite many schematization efforts, most 
services do not come with an explicit schema. At best, the documentation provides sample responses 
for typical requests.

For example, \url{http://openweathermap.org/current} provides an end-point to get the current weather.
The documentation shows one sample to illustrate the typical response. Using 
standard JSON and web libraries, we might write:

@@

\item The type providers prefer F\# object types with properties. This allows extensible 
  (open-world) data formats (\S\ref{sec:providers-xml}). It also interacts well with developer tooling,
  as most F\# editors provide auto-completion on ``.'' making values of provided types easy to use.
  This also makes our techniques applicable any language with nominal object types -- for example
  variations of Java or C# with a type provider mechanism added.

\item Finally, the mechanism handles practical concerns including
  support for different numerical types, \kvd{null} values and missing data.
\end{itemize}

@@

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter} 
to the type provider. Static parameters are resolved at compile-time and have to be constant. 
The provider analyzes the sample and provides a type  \ident{People}. F\# editors also 
execute the type provider in the background at development-time and use the provided types 
in auto-completion and background type-checking.

@@

The shape algebra and inference presented here is influenced by the design principles
we outlined earlier and the type definitions available in the F\# language.
If applied to other languages, details may differe, for example
w.r.t. numerical types and missing data.

@@

We distinguish between \emph{non-nullable shapes} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record names and record field names. In the rest of the
paper, we assume that record fields can be freely reordered:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
\end{array}
\end{equation*}

@@

\noindent
Next, we extend the core model (sufficient for the relativized safety) 
with \emph{labelled top types} discussed earleir.

>>>>

Applications for social networks, finding tomorrow's weather or searching train schedules
all communicate with external services. Increasingly, these services provide end-points that return 
data as CSV, XML or JSON. Most such services do not come with an explicit schema. At best, the 
documentation provides sample responses for typical requests.

For example, \url{http://openweathermap.org/current} documents an end-point to get the current 
weather with one example. Using standard libraries, we might call it as:

@@

\item The type providers prefer F\# object types with properties. This allows extensible 
  (open-world) data formats (\S\ref{sec:providers-xml}) and it interacts well with developer tooling
  (\S\ref{sec:providers-json}).

\item The above also makes our techniques applicable to any language with nominal 
  object types (e.g.~variations of Java or C\# with a type provider mechanism added).

\item Finally, the we handle practical concerns including
  support for different numerical types, \kvd{null} and missing data.
\end{itemize}

@@

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter} 
to the type provider. Static parameters are resolved at compile-time and have to be constant. 
The provider analyzes the sample and provides a type  \ident{People}. F\# editors also 
execute the type provider in the background at development-time and use the provided types 
in auto-completion on ``.'' and background type-checking.

@@

The shape algebra and inference presented here is influenced by the design principles
we outlined earlier and the type definitions available in the F\# language.
If applied to other languages, details may differe, for example
with respect to numerical types and missing data.

@@

We distinguish between \emph{non-nullable shapes} that always have a valid value (written as 
$\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
(written as $\sigma$). We write $\nu$ for record names and record field names. In the rest of the
paper, we assume that record fields can be freely reordered:

\noindent
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
\end{array}
\end{equation*}

@@

\noindent
Next, we extend the core model (sufficient for the relativized safety) 
with \emph{labelled top types} discussed earlier.
